Transfer of Value Functions via Variational Methods

### 2018-05-18 NIPS 2018/main.tex
aaa

\begin{restatable}{theorem}{thmg} \label{th:main-mixture}
Fix a target task $\mathcal{M}_{\tau}$. Assume linearly parameterized value functions $Q_{\bm{w}}(s,a)=\bm{w}^T\bm{\phi}(s,a)$ with bounded weights $\norm{\bm{w}}_2\leq w_{max}$ and uniformly bounded features $\norm{\bm{\phi}(s,a)}_2\leq \phi_{\max}$. Consider the mixture version of Algorithm~ \ref{alg:transfer} using $C$ components, source task weights $\mathcal{W}_s$, and bandwidth $\sigma^2_p\ $ for the prior. Denote by $\wh{\bm{\xi}} = (\wh{\bm{\mu}}_1,\dots,\wh{\bm{\mu}}_C,\wh{\bm{\Sigma}}_1,\dots,\wh{\bm{\Sigma}}_C)$ the variational parameters minimizing the objective of Eq. \eqref{eq:elbo} on a dataset $D$ of $N$ i.i.d. samples distributed according to $\tau$ and $\nu$. Let $\bm{w}^* = \arginf_{\bm{w}}||\wt{B}_{\bm{w}}||_{\nu}^2$ and define $\upsilon(\bm{w}^*) \triangleq \mathbb{E}_{\mathcal{N}(\bm{w}^*,\frac{1}{N}\bm{I})}\left[ \text{v}(\bm{w}) \right]$, with $\text{v}(\bm{w}) \triangleq \mathbb{E}_{\nu}\left[Var_{\mathcal{P}_{\tau}}\left[\wt b(\bm{w})\right]\right]$. Then, there exist constants $c_1,c_2,c_3$ such that, with probability at least $1-\delta$ over the choice of the dataset $D$:
\begin{equation}
\resizebox{0.93\textwidth}{!}{$
\mathbb{E}_{q_{\wh{\bm{\xi}}}}\left[ \norm{\wt B_{\bm{w}}}_{\nu}^2 \right ] \leq\  2\norm{\wt{B}_{\bm{w}^{*}}}_{\nu}^2 + \upsilon(\bm{w}^*) + c_1\sqrt{\frac{\log\frac{2}{\delta}}{N}} + \frac{c_2 + \lambda d \log N + 2\lambda\varphi\left(\Delta\right)}{N} + \frac{c_3}{N^2}$},
\end{equation}
where, $\Delta$ is a vector whose $j$-th components is $\Delta = \frac{1}{2\sigma^2_p}\norm{\bm{w}^* - \bm{w}_j}$ and, for a vector $\bm{x} = (x_1,\dots,x_d)$, $\varphi \left( \bm{x}\right) \triangleq \sum_i \frac{e^{-x_i}}{\sum_j e^{-x_j}}x_i$ is the \textit{softmin} function.
\end{restatable}

@@

	\begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[trim=0.0cm 0cm 1.2cm 1.3cm,clip=true,height=2.9cm]{images/cartpole/lrev.pdf}
    \caption{Cartpole}\label{f:cartpole}
  \end{subfigure}
    \begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[trim=0.0cm 0cm 1.2cm 1.3cm,clip=true,height=2.9cm]{images/mountaincar/lrev.pdf}
    \caption{Mountain Car}\label{f:mountaincar}
  \end{subfigure}
    \begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[trim=0.0cm 0cm 1.2cm 1.3cm,clip=true,height=2.9cm]{images/maze/lrev_main.pdf}
    \caption{Maze Navigation}\label{f:maze_results}
  \end{subfigure}
  \caption{Expected return as a function of the number of iterations averaged over $20$ independent runs. $95\%$ confidence intervals are shown.}
  \label{fig:main}
\end{figure*}
We consider an agent navigating in the environment depicted in Figure~\ref{fig:main}. The agent starts in the bottom-left corner and must move from one room to another to reach the goal position in the top-right corner. The rooms are connected by small doors whose locations are unknown to the agent. The state-space is modeled as a $10 \times 10$ continuous grid, while the action-space is the set of $4$ movement directions (up, right, down, left). After each action, the agent moves by $1$ in the chosen direction and the final position is corrupted by Gaussian noise $\mathcal{N}(0,0.2)$. In case the agent hits a wall, its position remains unchanged. The reward is $1$ when reaching the goal (after which the process terminates) and $0$ otherwise, while the discount factor is $\gamma = 0.99$. In this experiment, we consider linearly parameterized $Q$-functions with $121$ equally-spaced radial basis features.\\
We generate a set of $50$ source tasks for the three-room environment of Figure~\ref{fig:main} by sampling both door locations uniformly in the allowed space, and solve all of them by directly minimizing the TD error as presented in Section~\ref{sec:td}. Then, we use our algorithms to transfer from $10$ source tasks sampled from the previously generated set. The average return over the last $50$ learning episodes as a function of the number of iterations is shown in Figure~\ref{f:two_doors}. Each curve is the result of $20$ independent runs, each one resampling the target and source tasks, with $95\%$ confidence intervals. Further details on the parameters adopted in this experiment are given in Appendix \ref{app:c1}. As expected, the no-transfer (NT) algorithm fails at learning the task in so few iterations due to the limited exploration provided by an $\epsilon$-greedy policy. On the other hand, all our algorithms achieve a significant speed-up and converge to the optimal performance in few iterations, with GVT being slightly slower. Interestingly, we notice that there is no advantage in adopting more than $1$ component for the posterior in MGVT. This result is intuitive since, as soon as the algorithm figures out which is the target task, all the components move towards the same region.\\
To better understand the differences between GVT and MGVT, we now consider transferring from a slightly different distribution than the one from which target tasks are drawn. We generate $50$ source tasks again but this time with the bottom door fixed at the center and the other one moving. Then, we repeat the previous experiment, allowing both doors to move when sampling target tasks. The results are shown in Figure~\ref{f:one_door}. Interestingly, MGVT seems almost unaffected by this change, proving that it has sufficient representation power to generalize to slightly different task distributions. The same does not hold for GVT, which now is not able to solve many of the sampled target tasks, as can be noticed from the higher variance. This result proves again that assuming Gaussian distributions can pose severe limitations in our transfer settings.

@@

We presented a variational method for transferring value functions in RL. We showed our approach to be general, in the sense that it can be combined with several distributions and function approximators, while providing two practical algorithms based on Gaussians and mixtures of Gaussians, respectively. We analyzed both from a theoretical and empirical perspectives, proving that the Gaussian version has severe limitations, while the mixture one is much better for our transfer settings.

>>>>

\begin{restatable}{theorem}{thmg} \label{th:main-mixture}
Fix a target task $\mathcal{M}_{\tau}$. Assume linearly parameterized value functions $Q_{\bm{w}}(s,a)=\bm{w}^T\bm{\phi}(s,a)$ with bounded weights $\norm{\bm{w}}_2\leq w_{max}$ and uniformly bounded features $\norm{\bm{\phi}(s,a)}_2\leq \phi_{\max}$. Consider the mixture version of Algorithm~ \ref{alg:transfer} using $C$ components, source task weights $\mathcal{W}_s$, and bandwidth $\sigma^2_p\ $ for the prior. Denote by $\wh{\bm{\xi}} = (\wh{\bm{\mu}}_1,\dots,\wh{\bm{\mu}}_C,\wh{\bm{\Sigma}}_1,\dots,\wh{\bm{\Sigma}}_C)$ the variational parameters minimizing the objective of Eq. \eqref{eq:elbo} on a dataset $D$ of $N$ i.i.d. samples distributed according to $\tau$ and $\nu$. Let $\bm{w}^* = \arginf_{\bm{w}}||\wt{B}_{\bm{w}}||_{\nu}^2$ and define $\upsilon(\bm{w}^*) \triangleq \mathbb{E}_{\mathcal{N}(\bm{w}^*,\frac{1}{N}\bm{I})}\left[ \text{v}(\bm{w}) \right]$, with $\text{v}(\bm{w}) \triangleq \mathbb{E}_{\nu}\left[Var_{\mathcal{P}_{\tau}}\left[\wt b(\bm{w})\right]\right]$. Then, there exist constants $c_1,c_2,c_3$ such that, with probability at least $1-\delta$ over the choice of the dataset $D$:
\begin{equation}
\resizebox{0.93\textwidth}{!}{$
\mathbb{E}_{q_{\wh{\bm{\xi}}}}\left[ \norm{\wt B_{\bm{w}}}_{\nu}^2 \right ] \leq\  2\norm{\wt{B}_{\bm{w}^{*}}}_{\nu}^2 + \upsilon(\bm{w}^*) + c_1\sqrt{\frac{\log\frac{2}{\delta}}{N}} + \frac{c_2 + \lambda d \log N + 2\lambda\varphi\left(\Delta\right)}{N} + \frac{c_3}{N^2}$},
\end{equation}
where, $\Delta$ is a vector whose $j$-th component is $\Delta_j = \frac{1}{2\sigma^2_p}\norm{\bm{w}^* - \bm{w}_j}$ and, for a vector $\bm{x} = (x_1,\dots,x_d)$, $\varphi \left( \bm{x}\right) \triangleq \sum_i \frac{e^{-x_i}}{\sum_j e^{-x_j}}x_i$ is the \textit{softmin} function.
\end{restatable}

@@

	\begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[trim=0.0cm 0cm 1.2cm 1.3cm,clip=true,height=2.9cm]{images/cartpole/lrev.pdf}
    \caption{Cartpole}\label{f:cartpole}
  \end{subfigure}
    \begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[trim=0.0cm 0cm 1.2cm 1.3cm,clip=true,height=2.9cm]{images/mountaincar/lrev.pdf}
    \caption{Mountain Car}\label{f:mountaincar}
  \end{subfigure}
    \begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[trim=0.0cm 0cm 1.2cm 1.3cm,clip=true,height=2.9cm]{images/maze/lrev_main.pdf}
    \caption{Maze Navigation}\label{f:maze_results}
  \end{subfigure}
  \caption{Expected return as a function of the number of iterations averaged over $20$ independent runs. $95\%$ confidence intervals are shown.}
  \label{fig:main}
\end{figure*}
We consider an agent navigating in the environment depicted in Figure~\ref{fig:main}. The agent starts in the bottom-left corner and must move from one room to another to reach the goal position in the top-right corner. The rooms are connected by small doors whose locations are unknown to the agent. The state-space is modeled as a $10 \times 10$ continuous grid, while the action-space is the set of $4$ movement directions (up, right, down, left). After each action, the agent moves by $1$ in the chosen direction and the final position is corrupted by Gaussian noise $\mathcal{N}(0,0.2)$. In case the agent hits a wall, its position remains unchanged. The reward is $1$ when reaching the goal (after which the process terminates) and $0$ otherwise, while the discount factor is $\gamma = 0.99$. In this experiment, we consider linearly parameterized $Q$-functions with $121$ equally-spaced radial basis features.

We generate a set of $50$ source tasks for the three-room environment of Figure~\ref{fig:main} by sampling both door locations uniformly in the allowed space, and solve all of them by directly minimizing the TD error as presented in Section~\ref{sec:td}. Then, we use our algorithms to transfer from $10$ source tasks sampled from the previously generated set. The average return over the last $50$ learning episodes as a function of the number of iterations is shown in Figure~\ref{f:two_doors}. Each curve is the result of $20$ independent runs, each one resampling the target and source tasks, with $95\%$ confidence intervals. Further details on the parameters adopted in this experiment are given in Appendix \ref{app:c1}. As expected, the no-transfer (NT) algorithm fails at learning the task in so few iterations due to the limited exploration provided by an $\epsilon$-greedy policy. On the other hand, all our algorithms achieve a significant speed-up and converge to the optimal performance in few iterations, with GVT being slightly slower. Interestingly, we notice that there is no advantage in adopting more than $1$ component for the posterior in MGVT. This result is intuitive since, as soon as the algorithm figures out which is the target task, all the components move towards the same region.

To better understand the differences between GVT and MGVT, we now consider transferring from a slightly different distribution than the one from which target tasks are drawn. We generate $50$ source tasks again but this time with the bottom door fixed at the center and the other one moving. Then, we repeat the previous experiment, allowing both doors to move when sampling target tasks. The results are shown in Figure~\ref{f:one_door}. Interestingly, MGVT seems almost unaffected by this change, proving that it has sufficient representation power to generalize to slightly different task distributions. The same does not hold for GVT, which now is not able to solve many of the sampled target tasks, as can be noticed from the higher variance. This result proves again that assuming Gaussian distributions can pose severe limitations in our transfer settings.

@@

We presented a variational method for transferring value functions in RL. We showed our approach to be general, in the sense that it can be combined with several distributions and function approximators, while providing two practical algorithms based on Gaussians and mixtures of Gaussians, respectively. We analyzed both from a theoretical and empirical perspectives, proving that the Gaussian version has severe limitations, while the mixture one is much better for our transfer settings. We evaluated the proposed algorithms in different domains, showing that both achieve excellent performance in simple tasks, while only the mixture version is able to handle complex environments.
